---
title: "sotk2 quickstart: single-cohort PV example"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
author:
- name: Heewon Seo
abstract: >
    This quickstart walks through the core sotk2 workflow using the bundled
    Nanostring GeoMx DSP parvalbumin-interneuron (PV) dataset (one cohort,
    ranks 2 to 6). It builds a SOSet, constructs the metagene correlation
    network and detects communities (SOTK), and shows the corresponding
    network and rank-coverage plots. For the full multi-cohort demonstration
    (GLASS, IVYGAP, HEILAND), see the online documentation.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sotk2 quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Overview

This quickstart uses the bundled PV (parvalbumin interneuron) GeoMx DSP demo to show the minimum end-to-end workflow:

  * Build a `SOSet` from a single NMF.rank object across a chosen rank grid.
  * Compute the metagene-metagene correlation matrix.
  * Threshold the correlations, detect communities, and inspect the network.
  * Look at rank coverage to interpret which ranks contribute to which communities.

For the multi-cohort workflow (GLASS, IVYGAP, HEILAND) and the full annotation and signature-extraction examples, see <https://Snyder-Institute.github.io/sotk2/>.

## 0. Data preparation
Users can use either [NMF](https://github.com/renozao/NMF) or [cNMF](https://github.com/dylkot/cNMF) to deconvolute transcriptomics profiles to obtain metagenes/GEPs (gene expression programs).

### Run NMF
```{r runNMF, eval=FALSE, echo=TRUE}
# Do not Run
library(NMF)
library(Biobase)

## Test data
data(esGolub)
esGolub <- esGolub[c(1:200), c(1:20)]

## NMF options
method <- "brunet"; nr <- 10; seed <- 123456; options <- "tv2p4" # examples

## Run NMF
nmfRes <- nmf(esGolub, rank = c(2:5), method = method, nrun = nr, seed = seed, .options = options)
saveRDS(nmfRes, "NMF_rank_2_5.RDS")
```

### Run NMF independently for each rank and merge them into a single NMF object
```{r runNMF_parallel, eval=FALSE, echo=TRUE}
# Do not Run
library(NMF)
library(Biobase)

## Test data
data(esGolub)
esGolub <- esGolub[c(1:200), c(1:20)]

## NMF options
method <- "brunet"; nr <- 10; seed <- 123456; options <- "tv2p4" # examples

## Run NMF in parallel
### on Server A
res2 <- nmf(esGolub, rank=2, method=method, nrun=nr, seed=seed, .options=options)
saveRDS(res2, "NMF_rank_2.RDS")

### on Server B
res3 <- nmf(esGolub, rank=3, method=method, nrun=nr, seed=seed, .options=options)
saveRDS(res3, "NMF_rank_3.RDS")

### on Server C
res4 <- nmf(esGolub, rank=4, method=method, nrun=nr, seed=seed, .options=options)
saveRDS(res4, "NMF_rank_4.RDS")

## Load NMF results and combine them with sotk2::mergeNMFObjs()
library(sotk2)
resRank2 <- readRDS("NMF_rank_2.RDS")
resRank3 <- readRDS("NMF_rank_3.RDS")
resRank4 <- readRDS("NMF_rank_4.RDS")

nmfRes <- mergeNMFObjs(list(`2` = resRank2, `3` = resRank3, `4` = resRank4))
saveRDS(nmfRes, "NMF_outputs.RDS")
```

## 1. Quick start
The example below runs sotk2 on the Nanostring GeoMx DSP demo from mouse brain, where two segments (parvalbumin interneurons, PV+; and neurons, NeuN+) were profiled in the retrosplenial cortex. The bundled NMF.rank object is the PV segment across ranks 2 to 6.

### Load the package and parameter settings
```{r preparation}
library(sotk2)

## Load the NMF object
pvObjFile <- system.file(
        "extdata", "PV.RDS",
        package = "sotk2"
)
pvNmfObj <- readRDS(pvObjFile)

## Parameter settings
dataL <- list(PV = pvNmfObj)
rankL <- list(PV = c(2:6))
dataColor <- c("PV" = "#7570B3")
corMethod <- "spearman"

corrCoefThrehold <- 0.5
seed <- 1118
iteration <- 1000
communityWeight <- 100 # for visualization
cohortWeight <- 1 # tweak for multiple cohorts

nodeSize <- 10
nodeLabelSize <- 2

modelCols <- c(
        "TG1F" = "#084594",
        "TG2F" = "#2171B5",
        "TG3F" = "#4292C6",
        "TG4F" = "#6BAED6",
        "WT1F" = "#99000D",
        "WT2F" = "#CB181D",
        "WT3F" = "#EF3B2C"
)

```

### Concat W matrics and calculate pairwise correlations
```{r input_correlationMatrix}
soSet <- SOSet(NMFobjL = dataL, NMFrankL = rankL, dataCol = dataColor, corMet = corMethod)
plotCorrDensity(soSet, filename = "01_Stats_correlation_density.pdf", width = 6, height = 6)
```

### Generate a correlation network and apply community search
```{r run_sotk}
soObj <- SOTK(soSet, coefThre = corrCoefThrehold, seed = seed, niter = iteration, commWeight = communityWeight, cohortWeight = cohortWeight)
```

### Visualization
```{r network_visualization}
par(mfrow=c(1,1))
plotNetwork(soObj, label = FALSE, annot = "cohort", edgeAlpha = 0.8, weighted = FALSE, filename = "02_Network_Unweighted.pdf", vertexSize = nodeSize, vertexLabelCex = nodeLabelSize)
plotNetwork(soObj, label = TRUE, annot = "cohort", edgeAlpha = 0.8, weighted = FALSE, filename = "03_Network_Unweighted_lbl.pdf", vertexSize = nodeSize, vertexLabelCex = nodeLabelSize) 
plotNetwork(soObj, label = FALSE, annot = "community", edgeAlpha = 0.8, weighted = FALSE, filename = "04_Network_Unweighted_Comm.pdf", vertexSize = nodeSize, vertexLabelCex = nodeLabelSize)
plotNetwork(soObj, label = TRUE, annot = "community", edgeAlpha = 0.8, weighted = FALSE, filename = "05_Network_Unweighted_Comm_lbl.pdf", vertexSize = nodeSize, vertexLabelCex = nodeLabelSize)
plotNetwork(soObj, label = FALSE, annot = "community", edgeAlpha = 0.8, weighted = TRUE, filename = "06_Network_Community.pdf", vertexSize = nodeSize, vertexLabelCex = nodeLabelSize)
plotNetwork(soObj, label = TRUE, annot = "community", edgeAlpha = 0.8, weighted = TRUE, filename = "07_Network_Community_lbl.pdf", vertexSize = nodeSize, vertexLabelCex = nodeLabelSize)
plotNetwork(soObj, label = FALSE, annot = "cohort", edgeAlpha = 0.8, weighted = TRUE, filename = "08_Network_Weighted.pdf", vertexSize = nodeSize, vertexLabelCex = nodeLabelSize)
plotNetwork(soObj, label = TRUE, annot = "cohort", edgeAlpha = 0.8, weighted = TRUE, filename = "09_Network_Weighted_lbl.pdf", vertexSize = nodeSize, vertexLabelCex = nodeLabelSize)
```

### Sample annotation on the correlation network
```{r sample_annotation}
.getModelStats <- function(x) {
        TG1F <- 0; TG2F <- 0; TG3F <- 0; TG4F <- 0; 
        WT1F <- 0; WT2F <- 0; WT3F <- 0

        if (!is.null(x)) {
                for (i in x) {
                        buff <- unlist(stringr::str_split(i, "-"))
                        if (buff[1] == "TG1F") {
                                TG1F <- TG1F + 1
                        } else if (buff[1] == "TG2F") {
                                TG2F <- TG2F + 1
                        } else if (buff[1] == "TG3F") {
                                TG3F <- TG3F + 1
                        } else if (buff[1] == "TG4F") {
                                TG4F <- TG4F + 1
                        } else if (buff[1] == "WT1F") {
                                WT1F <- WT1F + 1
                        } else if (buff[1] == "WT2F") {
                                WT2F <- WT2F + 1
                        } else if (buff[1] == "WT3F") {
                                WT3F <- WT3F + 1
                        } else {
                                message(paste0("WARNING::", buff[1], " is uncategorized."))
                        }
                }
        }
        return(c(TG1F, TG2F, TG3F, TG4F,
                WT1F, WT2F, WT3F))
}

sample <- "TG4F"
graph <- soObj@corNetwork
layout <- soObj@unweightedLay # or @weightedLay
nodes <- igraph::V(graph)$name
clusterInfo <- soObj@sample2metagene

# Initialization
vertexPie <- as.list(rep(1, length(nodes))); names(vertexPie) <- nodes
vertexPieCol <- as.list(rep("white", length(nodes))); names(vertexPieCol) <- nodes
vertexSize <- rep(0, length(nodes)); names(vertexSize) <- nodes

for (ndx in seq_along(nodes)) {
        node <- nodes[ndx]
        samples <- clusterInfo[["PV"]][[node]]
        sdx <- which(sapply(stringr::str_split(samples, "-"), "[[", 1) == sample)

        if (length(samples) > 0 && length(sdx) > 0) {
                vertexSize[[node]] <- 5
                vertexPie[[node]] <- .getModelStats(unique(samples[sdx]))
                vertexPieCol[[node]] <- modelCols
        }
}

pdf(paste0("10_Model_Annotation_", sample, ".pdf"), width=10, height=10)
plot(graph,
        layout = layout,
        vertex.label = NA,
        vertex.size = as.numeric(vertexSize),
        vertex.shape = "pie",
        vertex.pie = vertexPie,
        vertex.pie.color = vertexPieCol,
        edge.color = scales::alpha("grey80", 0.3)
)
title(main = sample)
dev.off()
```

### Check the metagene coverage at each rank
```{r select_rank}
colComm <- soObj@commCols
ranks <- sort(as.numeric(unique(sapply(stringr::str_split(nodes, "\\$"), "[[", 3))))

pdf("11_coverage.pdf", width = 10, height = 10)
for (rank in ranks) {
        vertexLabel <- paste(
                sapply(stringr::str_split(nodes, "\\$"), "[[", 2),
                sapply(stringr::str_split(nodes, "\\$"), "[[", 3),
                sep = "/"
        )
        vertexLabel[which(as.numeric(sapply(stringr::str_split(nodes, "\\$"), "[[", 3)) != rank)] <- ""

        vertexColor <- colComm[igraph::V(graph)$community]
        liveCols <- unique(vertexColor[which(as.numeric(sapply(stringr::str_split(nodes, "\\$"), "[[", 3)) == rank)])
        if (length(liveCols) > 0) {
                vertexColor[-which(vertexColor %in% liveCols)] <- "white"
        }
        vertexColor[which(vertexLabel != "")] <- "black"
        liveColNode <- length(which(vertexColor == "black"))

        plot(graph,
                layout = layout,
                vertex.label = vertexLabel,
                vertex.color = vertexColor,
                vertex.label.color = "white",
                vertex.size = 8,
                vertex.label.cex = 0.75,
                edge.color = "grey80"
        )
        title(paste0("Rank = ", rank), cex.main = 2, col.main = "black")
        legend("bottomleft",
                legend = c(
                        paste0(liveColNode, " metagenes"),
                        paste0(length(colComm), " communities"),
                        paste0("Ratio = ", round(length(colComm) / liveColNode, 2))
                ),
                bty = "n"
        )
}
dev.off()
```
